---
title: "The Evolution of Esports Earnings: A Statistical Analysis of Prize Money, Tournaments, and Players (1998–2025)"
author:
  - name: "JiXianLi"
  - name: "Junbo Zhao"
  - name: "Name"
date: today
format:
  pdf:
    toc: false
    number-sections: true
    number-depth: 5
    fig-align: center
    cap-location: top
    geometry:
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
    colorlinks: true
execute:
  echo: false
  warning: false
---
```{r}
library(tidyverse)
library(ggplot2)
library(scales)
```

# Introduction
Esports has gotten a lot bigger over the last thirty years. What used to be a pretty niche hobby with small LAN tournaments has turned into an industry with millions of viewers and massive prize pools. The number of professional players and organized events keeps going up, and at this point competitive gaming is treated more like a real sport than it used to be. But even though esports is clearly growing, there is not a lot of detailed analysis on what that growth actually looks like when you dig into the numbers. We were curious about a few things in particular: whether the total prize money has been increasing steadily or in bursts, how evenly the money is spread across tournaments and players in a given year, and whether having more active players actually leads to bigger prize pools. To look into this, we used a dataset that covers esports from 1998 to 2025 and includes total prize money, tournament counts, mean and median prize pool per tournament, and mean and median earnings per player. The report is split into three main sections — we first look at how the overall prize pool has trended over time, then compare mean and median to see how skewed the distributions are, and finally explore the relationship between the number of active players and total prize money.

\newpage

# Data Analysis
## Total Prize Trend
```{r}
#type the code of graph there
```
Analysis of the graph narrative

\newpage

## Mean VS. Medium(JunBoZhao)
```{r}
#| label: fig-MeanVsMedian
#| fig-cap: "Mean and median over time, on a log scale. Left: prize pool per tournament. Right: earnings per active player."
#| fig-pos: H
#| fig-width: 9
#| fig-height: 4.2

dataTable <- read.csv("Esports_Stats.csv", check.names = FALSE)

mvmLong <- dataTable |>
  select(year,
         `Mean Tournament Prize Pool`,
         `Median Tournament Prize Pool`,
         `Mean Earnings/Player`,
         `Median Earnings/Player`) |>
  pivot_longer(-year, names_to = "metric", values_to = "value") |>
  mutate(
    panel  = ifelse(grepl("Tournament", metric),
                    "Per tournament prize pool",
                    "Per player earnings"),
    centerType = ifelse(grepl("^Mean", metric), "Mean", "Median")
  )

ggplot(mvmLong, aes(x = year, y = value, color = centerType)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  facet_wrap(~ panel, scales = "free_y") +
  scale_y_log10(labels = dollar) +
  scale_color_manual(values = c("Mean" = "#c0392b", "Median" = "#2c3e50")) +
  labs(
    title = "Where the average meets the typical",
    x = "Year", y = "USD (log scale)", color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold"))
```

```{r}
#| label: fig-MMratio
#| fig-cap: "Ratio of mean to median over time. A value of 1 (dashed line) would mean a perfectly balanced distribution; everything above it is right-skew."
#| fig-pos: H
#| fig-width: 9
#| fig-height: 3.5

dataTable <- read.csv("Esports_Stats.csv", check.names = FALSE)

ratioLong <- dataTable |>
  transmute(
    year,
    `Tournament prize pool` =
      `Mean Tournament Prize Pool` / `Median Tournament Prize Pool`,
    `Per-player earnings`   =
      `Mean Earnings/Player`     / `Median Earnings/Player`
  ) |>
  pivot_longer(-year, names_to = "metric", values_to = "ratio")

ggplot(ratioLong, aes(x = year, y = ratio, color = metric)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  scale_color_manual(values = c("Tournament prize pool" = "#3b7a57",
                                "Per-player earnings"   = "#d68c45")) +
  labs(
    title = "How skewed was each year?",
    x = "Year", y = "Mean / Median", color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

We wanted to compare the mean and median here because they pick up on different things. The mean gets pulled around by outliers — one massive tournament like The International will push it way up even if most events that year were small. The median does not care about that at all; it just shows what the middle tournament looks like. So when these two numbers are far apart, it usually means the distribution is really skewed.

Looking at @fig-MeanVsMedian, we had to use a log scale because the range is kind of ridiculous. In 2010 the median tournament was paying around \$250, but the mean was already over \$6,000. On a normal axis you would not even be able to see the median line. From about 2010 to 2021, the two lines on the tournament side barely look related — the median stays somewhere in the low hundreds to low thousands while the mean is up in the tens of thousands. We plotted the actual ratio in @fig-MMratio to get a clearer picture, and for most of that stretch it is somewhere between 12x and 29x. Basically, for every big event like LoL Worlds there were a ton of small local tournaments that only paid out a couple hundred dollars, and those kept dragging the median down.

On the player earnings side, the gap is there too but not as extreme. The mean is usually around 3 to 7 times the median, with a peak at about 15x in 2014. So a handful of top players were making a lot, but most people competing were earning way less than the average would suggest.

What caught our attention was the shift after 2022. The median tournament prize pool went from roughly \$1,500 in 2021 up to over \$18,000 by 2025, and the mean-to-median ratio dropped from around 26 all the way to about 3. At first that looks like a good thing, like the money is spreading out more evenly. But when we checked the total number of tournaments, the count went from about 6,300 in 2019 down to around 1,700 in 2025. So what actually happened is a lot of the smaller grassroots events stopped running. The tournaments that are left tend to be bigger and better funded, which is why the median jumped up and closed the gap with the mean. It is less that the distribution got more equal and more that the bottom end just went away.

\newpage

## The Relationship Between The Number Of Players And The Award Pool Of The Tournament(JiXianLi)
```{r}
#| label: fig-PlayersAndPrice
#| fig-cap: "Relationship Between Total Players and Award Pool Of Tournament"
#| fig-pos: H
#| fig-width: 9
#| fig-height: 4.5

dataTable <- read.csv("Esports_Stats.csv", check.names = FALSE)
ggplot(dataTable, aes(x = `Total Active Players`, y = `Total Prize Money`)) +
  geom_line(color = "skyblue", linewidth = 1) +
  geom_point(color = "darkblue", size = 2) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Relationship Between Total Players and Award Pool Of Tournament",
    x = "Total Active Players",
    y = "Total Prize Money Of The Tournament (USD)"
  )
```
As the @fig-PlayersAndPrice shown, overall, there is a strong positive correlation.That is, the more active players there are, the higher the total prize money for the event. This is intuitive, as an increase in participants usually means an expansion of the market size, which attracts more sponsors, viewers and capital inflows, thereby raising the prize pool. Moreover, this line is not a uniform straight line, it can be split into three parts. First is player number from 0 to 10000, in this case the growth of bonuses has been very slow, which is in line with the early stage of the e-sports industry. At that time, the player base size was small and the commercialization level was low. Second, is player number from 10001 to 22000, in this case the slope has significantly steepened, and the bonus amount has rapidly increased from several hundred million dollars to nearly 200 million dollars. This indicates that within this range, the increase in bonus for each additional active player is much higher than in the early stages. Finally, is player number greater than 22001. An interesting phenomenon emerges here: the number of players decreases first and then rises sharply. This indicates that the relationship between the number of players and the prize money is no longer strictly linear. In some years, there are more players but the prize money is actually lower. Based on the observation, this "abnormal" period was from 2020 to 2021, which was the period of the COVID-19 pandemic. It makes sense because during the pandemic, a lot of people are staying at home, especially Chinese people (due to the policy). People staying at home may use video games to occupy their leisure time, but the tournaments cannot able to be held.

\newpage

# Summary
Analysis of the graph narrative

\newpage

# Code Appendix
```{r}
#| label: codeAppend
#| ref-label: !expr knitr::all_labels()
#| echo: true
#| eval: false
```
