---
title: "Whose game is it? Ten years of esports earnings"
author:
  - name: "[Name 1]"
  - name: "JiXianLi"
date: today
format:
  pdf:
    documentclass: article
    fontsize: 11pt
    geometry: margin=1in
    number-sections: true
    fig-align: center
    fig-pos: "H"
    keep-tex: false
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
csl: apa7.csl
---

```{r}
#| label: setup
#| include: false
# Style guide: Tidyverse (https://style.tidyverse.org/)
# Primary author: [Name 1]
# Reviewed by:    [Name 2]

library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(gt)
library(stringr)
library(here)

# helpers live in R/ so we don't clutter the report body
source(here("R", "scrape_helpers.R"))

# pre-tidied data loaded from data/ (produced by data-raw scripts)
games_by_year <- read_csv(here("data", "games_by_year.csv"),
                          show_col_types = FALSE)
```

# Introduction {#sec-intro}

Competitive video gaming has gone from LAN parties in basements to
stadium finals with eight-figure prize pools in about fifteen years.
Our team wanted to see what that trajectory actually looks like once
you pull the receipts: which titles have pulled the biggest cheques,
how that breakdown has moved between genres, and whether the money
has tracked the games people are actually playing.

Two of us follow esports casually — one of us has been watching the
International since 2015, the other got pulled in through Valorant
— but we're not going to lean on that. We'll treat the data as the
primary source of truth and keep our priors in check.

# Data and provenance {#sec-data}

## Primary source

Our main data came from the yearly "Top Games" pages on
`esportsearnings.com`, pulled for 2014 through 2024. The site is a
community-maintained database. Tournament results are submitted and
reviewed by volunteer editors; prize payouts get cross-checked against
official brackets and organizer announcements where available. A short
description of the editorial policy is linked in the site footer
[@esportsearnings_about].

We scraped with `{rvest}` wrapped in `{polite}` so the session respects
`robots.txt` and throttles to one request every few seconds. Raw HTML
is cached under `data-raw/cache/` so the scrape is reproducible
without repeatedly hitting the server. See the code appendix for the
full scraper.

## Supplementary source

To connect earnings to playerbase we pulled monthly peak concurrent
players from SteamDB / SteamCharts via their public endpoints. Only
PC titles that have a Steam page got matched; the remainder (console
fighters, mobile-only esports) are flagged and handled separately in
our analysis.

## Attributes we focus on

| Attribute | Type | Source | Notes |
|-----------|------|--------|-------|
| `year` | Quantitative (discrete) | esportsearnings | 2014–2024 |
| `game` | Categorical | esportsearnings | Title as listed on the site |
| `genre` | Categorical | Manual coding | MOBA, FPS, BR, Fighting, Card, Other |
| `total_earnings_usd` | Quantitative | esportsearnings | Sum of tournament prize money that year |
| `tournaments` | Quantitative | esportsearnings | Count of ranked events that year |
| `active_players` | Quantitative | esportsearnings | Distinct players earning prize money |
| `release_year` | Quantitative | esportsearnings game page | Used to compute game age |
| `steam_peak_ccu` | Quantitative | SteamCharts | Only for Steam-listed titles |

`genre` and `game_age` are attributes we derived, everything else comes
straight from the source.

## FAIR and CARE

**FAIR.** The data is Findable — esportsearnings has stable URLs per
year and per game. Accessibility is OK through the public site, but
there is no official API, so programmatic access depends on scraping
that might break if the site's layout changes. Interoperability is
limited: the data is published as HTML tables rather than a documented
schema. Reusability is the weakest link — the site doesn't publish a
formal license, and the terms of use allow personal and academic
reference but restrict bulk reproduction, which is why we cite heavily
and archive only what we need.

**CARE.** The data describes individuals (players' names, nationalities,
tournament winnings). None of that was collected with explicit
informed consent from the players themselves, since it originates from
public tournament records. We treat player-level attributes cautiously
— we aggregate to country level wherever we can, and we do not build
any kind of ranking of specific named individuals. Community benefit
and ethics are the dimensions we're most aware of here: the platform
benefits the esports community at large, but individual players are
the labour behind the numbers and deserve that framing.

# Exploratory data analysis {#sec-eda}

## Shape of the top-games table

```{r}
#| label: tbl-overview
#| tbl-cap: "Rows per year in the top-games table after tidying."
#| echo: false
games_by_year |>
  count(year, name = "games") |>
  gt() |>
  cols_label(year = "Year", games = "Games with recorded earnings") |>
  fmt_number(columns = games, decimals = 0)
```

<!-- Each team member is responsible for at least one table and one
     plot. We'll label authorship with the Code Header above each
     chunk once the analysis is in place. -->

## Prize money over time

```{r}
#| label: fig-total-by-year
#| fig-cap: "Total reported prize money across all top games, by year."
#| fig-alt: "A line plot showing total annual esports prize money climbing from roughly 30 million USD in 2014 to over 200 million in 2019, then falling during the pandemic and recovering afterwards."
#| fig-width: 6
#| fig-height: 3.5
games_by_year |>
  group_by(year) |>
  summarise(total_usd = sum(total_earnings_usd, na.rm = TRUE),
            .groups = "drop") |>
  ggplot(aes(x = year, y = total_usd / 1e6)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.5) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  labs(
    x = "Year",
    y = "Total prize money (USD millions)",
    caption = "Source: esportsearnings.com"
  ) +
  theme_minimal(base_size = 10)
```

*(placeholder narrative)* The rough shape is easy to describe: the
industry climbs steadily from 2014, spikes around 2018–2019 when
Fortnite brought in huge one-off prize pools, and then gets disrupted
by the pandemic before settling into a new baseline.

## Genre mix

*(to fill in once genre coding is done — the plan is a stacked area
chart with MOBA, FPS, Battle Royale, Fighting, Card, Other.)*

## Concentration of earnings

*(Lorenz-style cumulative plot of share of prize money across the top
N games, by year. Will also compute a simple Gini-like concentration
index and report it in a small table.)*

# Discussion {#sec-discussion}

*(Written once the plots are in. Keep the claims honest — this is
EDA, not inference.)*

# Author contribution {#sec-authors}

We follow the [CRediT taxonomy](https://credit.niso.org/) here.

- **[Name 1]:** Conceptualization, Data curation (scraping and
  tidying), Investigation, Visualization (time series), Writing –
  original draft.
- **[Name 2]:** Conceptualization, Data curation (Steam match),
  Formal analysis (concentration measures), Visualization (tables),
  Writing – review & editing.

# References {.unnumbered}

<!-- BibTeX entries live in references.bib -->

::: {#refs}
:::

# Code appendix {.unnumbered}

```{r}
#| label: appendix
#| echo: true
#| eval: false

# -- setup chunk --------------------------------------------------
# (see top of file)

# -- scraping -----------------------------------------------------
# see data-raw/01_scrape_esports.R

# -- tidying ------------------------------------------------------
# TODO: move the tidy steps here once finalized
```
