---
title: "Ten years of esports earning, how does the total bonus change over year!"
author:
  - name: "JiXianLi"
  - name: "Name"
  - name: "Name"
date: today
format: 
  pdf: 
    toc: false
    number-sections: true
    number-depth: 5
    fig-align: center
    cap-location: top
    geometry: 
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
    colorlinks: true
execute: 
  echo: false
  warning: false
---
```{r}
library(tidyverse)
library(ggplot2)
```

# Introduction
type the introduction here

\newpage

# Data Analysis
## Total Prize Trend
```{r}
#type the code of graph there
```
Analysis of the graph narrative

\newpage

## Mean VS. Medium
```{r}
#type the code of graph there
```
Analysis of the graph narrative

\newpage

## The Relationship Between The Number Of Players And The Prize Money Of The Tournament
```{r}
#type the code of graph there
```
Analysis of the graph narrative

\newpage

## Summary
Analysis of the graph narrative

\newpage

# Code Appendix
```{r}
#| label: codeAppend
#| ref-label: !expr knitr::all_labels()
#| echo: true
#| eval: false
```



