---
title: "The Evolution of Esports Earnings: A Statistical Analysis of Prize Money, Tournaments, and Players (1998–2025)"
author:
  - name: "JiXianLi"
  - name: "Name"
  - name: "Name"
date: today
format: 
  pdf: 
    toc: false
    number-sections: true
    number-depth: 5
    fig-align: center
    cap-location: top
    geometry: 
      - top=1in
      - left=1in
      - right=1in
      - bottom=1in
    colorlinks: true
execute: 
  echo: false
  warning: false
---
```{r}
library(tidyverse)
library(ggplot2)
library(scales)
```

# Introduction
E-sports has evolved from a small group of enthusiasts' entertainment into a global industry over the past three decades. At the same time, the number of professional players, the frequency of the competition, and the number of players are continuously increasing, which leads to esports becoming a formal exercise. However, we still lack quantitative research on the long-term growth model of e-sports. For instance, over the past nearly three decades, how has the total prize money of e-sports changed over time? Are there significant differences in the distribution of individual player incomes between different years? And is there a clear correlation between the growth of active players and the total prize money of. In order to answer these questions, this report records the data of e-sports from 1998 to 2025, including total award money, total number of tournaments, mean and median tournament award pool, and mean and median earnings per player. events? This report's analysis part mainly includes three parts: the trend of the award pool of the tournaments over the years, the difference between the mean and the median, and the relationship between the number of players and the award pool of the tournament.

\newpage

# Data Analysis
## Total Prize Trend
```{r}
#type the code of graph there
```
Analysis of the graph narrative

\newpage

## Mean VS. Medium(JunBoZhao)
```{r}
#type the code of graph there
```
Analysis of the graph narrative

\newpage

## The Relationship Between The Number Of Players And The Award Pool Of The Tournament(JiXianLi)
```{r}
#| label: fig-PlayersAndPrice
#| fig-cap: "Relationship Between Total Players and Award Pool Of Tournament"
#| fig-pos: H
#| fig-width: 9
#| fig-height: 4.5

dataTable <- read.csv("Esports_Stats.csv", check.names = FALSE)
ggplot(dataTable, aes(x = `Total Active Players`, y = `Total Prize Money`)) +
  geom_line(color = "skyblue", linewidth = 1) +
  geom_point(color = "darkblue", size = 2) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Relationship Between Total Players and Award Pool Of Tournament",
    x = "Total Active Players",
    y = "Total Prize Money Of The Tournament (USD)"
  )
```
As the @fig-PlayersAndPrice shown, overall, there is a strong positive correlation.That is, the more active players there are, the higher the total prize money for the event. This is intuitive, as an increase in participants usually means an expansion of the market size, which attracts more sponsors, viewers and capital inflows, thereby raising the prize pool. Moreover, this line is not a uniform straight line, it can be split into three parts. First is player number from 0 to 10000, in this case the growth of bonuses has been very slow, which is in line with the early stage of the e-sports industry. At that time, the player base size was small and the commercialization level was low. Second, is player number from 10001 to 22000, in this case the slope has significantly steepened, and the bonus amount has rapidly increased from several hundred million dollars to nearly 200 million dollars. This indicates that within this range, the increase in bonus for each additional active player is much higher than in the early stages. Finally, is player number greater than 22001. An interesting phenomenon emerges here: the number of players decreases first and then rises sharply. This indicates that the relationship between the number of players and the prize money is no longer strictly linear. In some years, there are more players but the prize money is actually lower. Based on the observation, this "abnormal" period was from 2020 to 2021, which was the period of the COVID-19 pandemic. It makes sense because during the pandemic, a lot of people are staying at home, especially Chinese people (due to the policy). People staying at home may use video games to occupy their leisure time, but the tournaments cannot able to be held.

\newpage

# Summary
Analysis of the graph narrative

\newpage

# Code Appendix
```{r}
#| label: codeAppend
#| ref-label: !expr knitr::all_labels()
#| echo: true
#| eval: false
```



